ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-27 pyked.tex
cleaned up a few things

%====================================================================
\section{Introduction}
%
Fundamental combustion experiments provide vital data for understanding fuel
chemistry and validating chemical kinetic models. Important measured quantities
include autoignition delay times, laminar flame speeds, and species profiles,
among others. However, despite both the importance and abundance of such
information in the literature, the combustion\slash chemical kinetics community
lacks an accepted, commonly used standard for recording and sharing data from
fundamental combustion experiments.\footnote{ Interestingly, this contrasts the
situation for chemical kinetic models, where the CHEMKIN~\autocite{Kee:1996ck}
format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or
FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code} lag behind considerably,
although use of the former continually grows due to Cantera's open availability.
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?
}

Instead, such data is typically found in under-documented comma-separated value
files and Excel spreadsheets, or contained in PDF tables rather than as
supplementary material associated with a paper. In the worst case, data is only
available in figures and must be digitized by the use of software such as
WebPlotDigitizer~\autocite{WebPlotDigitizer}. These practices limit the wide use
of the valuable, extensive experimental data available in the literature.
Frenklach~\autocite{Frenklach:2007bm} further explained the benefits of
standardized, widely available data for combustion. In brief, much of
fundamental combustion\slash chemical kinetics research drives towards the
ultimate goal of developing predictive kinetic models---and ultimately this
depends on a \emph{community} infrastructure for data and methods.

@@

Despite the development of these standards, limited examples of open data
\emph{sharing} utilizing the formats is available in the literature. To the best
of our knowledge, the largest such such publicly available database is hosted by
the Clean Combustion Research Center at King Abdullah University of Science and
Technology (KAUST) and is known as CloudFlame
(\url{https://cloudflame.kaust.edu.sa}). Started in
2013~\autocite{Goteng:2013cf},\todo{a few more refs needed}{} CloudFlame serves
as an openly accessible database for experimental data, available in a standard
CSV format, and also provides a cloud infrastructure for running simulations
based on stored models and data. While an admirable and useful effort,
CloudFlame, like ReSpecTh and PrIMe, is a closed system controlled by a single
institution rather than by the community at large.

Therefore, we believe there is a need for an open, community focused and
developed, combustion\slash chemical kinetics data format. In this work, we
present a new, open-source, human- and machine-readable data standard for
fundamental combustion experiments, \ck{}, and offer tools for easily working
with data encoded in this format. Niemeyer recently introduced an initial
version of the \ck{} format~\autocite{Niemeyer:2016wf}, and in this work we
further formalize and develop the standard. We also discuss a related Python
software tool for validating and working with \ck{} files: \pk{}. Our
motivations are similar those of the PrIMe and ReSpecTh teams: the \ck{} project
will enable easy sharing and use of fundamental combustion data, for the
(primary) purposes of developing and validating predictive chemical kinetic
models. However, we make usability a major design focus, and plan to share all
data and software openly to cultivate a community of user-contributors.

@@

The \yabox{value} can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox{author} mapping is used for any fields where
an author must be specified. The \yabox{author} mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox{name} (required, string): The author's full name
    \item \yabox{ORCID} (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
\todo[author=Kyle,inline]{one difference ReSpecTh called out from PrIMe is that
the latter added a copyright line in all files, claiming copyright on the data...
do we want to be incendiary and point out that you can't copyright data?}
%
\begin{itemize}
    \item \yabox{file-author} (required, \yabox{author}-type mapping): The
    author of the \ck{} file
    \item \yabox{file-version} (required, integer): The version of the \ck{}
    file
    \item \yabox{chemked-version} (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox{reference} (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox{authors} (required, sequence): A sequence of \yabox{author}
        mappings
        \item \yabox{journal} (required, string): The name of the publishing
        journal
        \item \yabox{year} (required, integer): The year of publication
    \end{itemize}
    \item \yabox{experiment-type} (required, string): Currently, only
    \yabox{ignition delay} is supported
    \item \yabox{apparatus} (required, string): The type of apparatus used to
    perform the experiment
\end{itemize}

@@

\ck{} and \pk{} are currently under active development, and contributions from
the community are welcomed. All development occurs under the BSD
3-clause license and the code is housed on GitHub at
\url{https://github.com/pr-omethe-us/PyKED}. Future directions for development
are outlined in the public
\href{Roadmap}{https://github.com/pr-omethe-us/PyKED/wiki/Roadmap}. The highest
priority issues are currently adding support for other types of fundamental
experiments, including flame speed, speciation, and flame extinction
measurements.

>>>>

%====================================================================
\section{Introduction}
%
Fundamental combustion experiments provide vital data for understanding fuel
chemistry and validating chemical kinetic models. Important measured quantities
include autoignition delay times, laminar flame speeds, and species profiles,
among others. However, despite both the importance and abundance of such
information in the literature, the combustion\slash chemical kinetics community
lacks an accepted, commonly used standard for recording and sharing data from
fundamental combustion experiments.\footnote{Interestingly, this contrasts the
situation for chemical kinetic models, where the CHEMKIN~\autocite{Kee:1996ck}
format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or
FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code} lag behind considerably,
although use of the former continually grows due to Cantera's open availability.
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?
}

Instead, such data is typically found in under-documented comma-separated value (CSV)
files and Excel spreadsheets, or contained in PDF tables rather than as
supplementary material associated with a paper. In the worst case, data is only
available in figures and must be digitized by the use of software such as
WebPlotDigitizer~\autocite{WebPlotDigitizer}. These practices limit the wide use
of the valuable, extensive experimental data available in the literature.
Frenklach~\autocite{Frenklach:2007bm} further explained the benefits of
standardized, widely available data for combustion. In brief, much of
fundamental combustion\slash chemical kinetics research drives towards the
ultimate goal of developing predictive kinetic models---and ultimately this
depends on a \emph{community} infrastructure for data and methods.

@@

Despite the development of these standards, limited examples of open data
\emph{sharing} using the formats exist in the literature. To the best
of our knowledge, the largest such publicly available database is hosted by
the Clean Combustion Research Center at King Abdullah University of Science and
Technology (KAUST) and is known as CloudFlame
(\url{https://cloudflame.kaust.edu.sa}). Started in
2013~\autocite{Goteng:2013cf},\todo{a few more refs needed}{} CloudFlame serves
as an openly accessible database for experimental data, available in a standard
CSV format, and also provides a cloud infrastructure for running simulations
based on stored models and data. While an admirable and useful effort,
CloudFlame, like ReSpecTh and PrIMe, is a closed system controlled by a single
institution rather than by the community at large.

Therefore, we believe there remains a need for an open, community-focused and
developed, combustion\slash chemical kinetics data format. In this work, we
present a new, open-source, human- and machine-readable data standard for
fundamental combustion experiments, \ck{}, and offer tools for easily working
with data encoded in this format. Niemeyer recently introduced an initial
version of the \ck{} format~\autocite{Niemeyer:2016wf}, and in this work we
further formalize and develop the standard. We also discuss a related Python
software tool for validating and working with \ck{} files: \pk{}. Our
motivations are similar those of the PrIMe and ReSpecTh teams: the \ck{} project
will enable easy sharing and use of fundamental combustion data, for the
(primary) purposes of developing and validating predictive chemical kinetic
models. However, we make usability a major design focus, and plan to share all
data and software openly to cultivate a community of user-contributors.

@@

The \yabox{value} can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox{author} mapping is used for any fields where
an author must be specified. The \yabox{author} mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox{name} (required, string): The author's full name
    \item \yabox{ORCID} (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
%
\begin{itemize}
    \item \yabox{file-author} (required, \yabox{author}-type mapping): The
    author of the \ck{} file
    \item \yabox{file-version} (required, integer): The version of the \ck{}
    file
    \item \yabox{chemked-version} (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox{reference} (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox{authors} (required, sequence): A sequence of \yabox{author}
        mappings
        \item \yabox{journal} (required, string): The name of the publishing
        journal
        \item \yabox{year} (required, integer): The year of publication
    \end{itemize}
    \item \yabox{experiment-type} (required, string): Currently, only
    \yabox{ignition delay} is supported
    \item \yabox{apparatus} (required, string): The type of apparatus used to
    perform the experiment
\end{itemize}

@@

\ck{} and \pk{} are currently under active development, and contributions from
the community are welcomed. All development occurs under the BSD
3-clause license and the code is housed on GitHub at
\url{https://github.com/pr-omethe-us/PyKED}. Future directions for development
are outlined in the public roadmap\footnote{\url{https://github.com/pr-omethe-us/PyKED/wiki/Roadmap}}.
The highest priority issues are currently adding support for other types of fundamental
experiments, including flame speed, speciation, and flame extinction
measurements.
